Prefix-free graphs and suffix array construction in sublinear space

### 2023-06-24 manuscript/conclusion.tex
finished sa construction

\section{Suffix array construction}
A suffix array is a permutation of string positions which lexicographically
sorts the suffixes of the string starting at that position.
It is an influential data structure with many applications in efficient string algorithms solving problems such as exact pattern matching, repeat finding, maximum exact match (MEM) finding, document retrieval and many more.

There exist several algorithms for suffix array construction in linear time \cite{skew2003karkkainen,sais2009nong,saoverview2007puglisi} with several practical implementations \cite{louza2017inducing,libdivsufsort}.
Despite their linear time complexity, these algorithms become bottlenecks in some applications because of their linear space complexity.
This observation is especially relevant in pangenomics, where the datasets often do not fit in the computer memory.

Here, we show another crucial advantage of prefix-free graphs.
Although they do not offer any improvement in theoretical guarantees in the worst case, in practice, they often represent the pangenome in a substantially smaller space and allow us to generate the suffix array values one by one, possibly using them directly in subsequential computation or storing them in compressed form.
This iteration can be done without ever expanding the pangenome to its textual representation in space proportional to the sum of segment lengths and the sum of path lengths.

\subsection{Iterator preparation}
To prepare the iterator of a suffix array of the pangenome from a prefix-free graph, we need to create several data structures.
First, we concatenate all the segments into a single string using a separator $\#$ and append a sentinel $\$$.
We will call this concatenation \emph{segment join}.
An example of a segment join is in Figure \ref{fig:ids_and_positions}.

Next, we calculate the segment join's suffix array and the longest common prefix (LCP) array.
For both of these arrays, there exist algorithms with linear time and space complexity which we can use.
We note that these linear complexities are proportional to the length of the segment join, which is usually much smaller than the original pangenome.

Next, for each suffix of the segment join, we need to calculate the corresponding segment ID and position values.
The value segment ID represents in what segment the current suffix starts, and the value segment position represents at what position in that particular segment the current suffix starts.
These arrays can be computed using an inverse permutation of a suffix array of a segment join in linear time.
To illustrate the procedure, consider the segment join of our running example from Figure \ref{fig:ids_and_positions}.
Each position of the join can be assigned a segment ID and a position in the current segment by linearly scanning the segment join and incrementing the ID and position accordingly.
Then, applying the inverse permutation of a suffix array to these arrays changes the order of computed values in correspondence to the sorted suffixes.
Table \ref{tab:suffix} shows the resulting suffix array, LCP array,
segment ID array and segment position array.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/ids_and_positions.png}
    \caption{
        Segment IDs and segment positions of a segment join. 
        Permuting these according to the inverse permutation of a suffix array results in segment ID array and segment position array shown in Table \ref{tab:suffix}.
    }
    \label{fig:ids_and_positions}
\end{figure}

\input{sa_table.tex}

In Table \ref{tab:suffix}, one row can represent multiple positions of the
pangenome.
To identify these positions, we store some additional information in Table \ref{tab:segment}.
For each segment, we store its length, starting positions in the pangenome and the rank of the right context of these positions.
To calculate the starting positions and the ranks of the right contexts, we use a \emph{path join}.
Similarly to segment join, a path join is a concatenation with delimiters $\#$ and sentinel $\$$, but now constructed by concatenating the paths.
An example of a path join for our running example is in Figure \ref{fig:path_join}.
Starting positions can be calculated by cumulatively summing the lengths
of the segments in path join and subtracting the overlaps.

The computation of ranks is more involved.
It uses the normalized form of prefix-free graphs since it relies on a lexicographically smaller ID in a path representing a lexicographically smaller segment.
We construct the suffix array of the path join and find its inverse permutation $\texttt{ISA}$.
$\texttt{ISA}$ gives us ranks for each position in the path join.
To determine the rank of the right context for position $i$, we take
the value of $\texttt{ISA}[i+1]$.
Finally, we store the starts and ranks sorted by the rank values in the segment table as shown in Table \ref{tab:segment}.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/path_join.png}
    \caption{}
    \label{fig:path_join}
\end{figure}

\input{segment_table.tex}

\subsection{Iteration}
With the previous tables stored in memory, we have all the necessary ingredients to generate the suffix array entries.

Each row in the suffix table represents a suffix of a particular segment.
There are four cases of what the first position of these suffixes can represent within the segments:
\begin{itemize}
    \item the sentinel
    \item a separator
    \item a position within the last $k$ characters of a segment
    \item a position outside the last $k$ characters, separator and sentinel
\end{itemize}

Since the original pangenome has no corresponding position for the sentinel or separator characters, we can skip the first rows representing them.

In the third case, the position is inside the trigger word or the sentinels appended during the graph construction.
The positions inside the trigger words are represented twice in the suffix table, once at the end of a segment and a second time at the beginning of the following segment in the pangenome.
These ending positions can violate the prefix-free property of the segment suffixes and, therefore, can be sorted incorrectly.
Skipping through these positions ensures the prefix-free property for the rest of the suffixes and also avoids double reporting.
This choice also plays nicely with the previous choice of appending $k$ sentinels during the construction of a prefix-free graph, as these positions will not get reported either.
In conclusion, if the length of a current segment suffix is smaller or equal to the size of the trigger words $k$, we skip the row as in the previous cases.

% stoped here
Finaly, in the last case we report the positions.
The last case will partition the suffix table into the blocks of the same suffixes.
As an example consider the rows 20 and 21, which together for a single block.
All other blocks in the example consist of single rows and therefore we call them singletons.

This partitioning leads to three cases (from simplest to hardest):
\begin{itemize}
    \item A singleton block, occuring only once in the whole pangenome
    \item A singleton block, repeating several times in the pangenome
    \item A non-singleton block
\end{itemize}

In the case of a singleton block with only a single occurence we need to report only a single suffix array entry.
Given the row index $i$, this entry can be calculated with the following equation:

\begin{equation}
    \label{eq:sa_entry}
    \texttt{SA entry} = \texttt{starts}[\texttt{ID}[i]] + \texttt{pos}[i]
\end{equation}

As an example, consider the row $13$ in the Suffix Table (\ref{tab:suffix}), the first row yielding a SA value.
It's segment ID is $0$ and from the Segment Table (\ref{tab:segment}) we see that there is only one occurence of segment $0$ and it starts at position $9$ in the pangenome.
The offset from the start of a segment $\texttt{pos}[13]$ is $0$.
This yields the first entry of a suffix array $9 + 0 = 9$ corresponding to the lexicographically smallest suffix $P[9..] = \texttt{ACACT}$.

The second, slightly more complex case is a singleton block with several occurences in the pangenome.
For this case, we need to report as many suffix array entries as the number of occurences.
Because the starting segment positions in the Segment Table (\ref{tab:segment}) are sorted based on their right context rank, we can iterate through these starting positions and apply Equation \ref{eq:sa_entry} to each of them.

As an example, consider the row $14$ in the Suffix Table (\ref{tab:suffix}).
This suffix occurs twice in the pangenome in segments starting at positions $15$ and $1$.
Since the offset from the start of a segment $\texttt{pos}[14]$ is $0$, we report a suffix array entry $15 + 0 = 15$ and $1 + 0 = 1$, corresponding to the suffixes $P[15..] = \texttt{ACGACT}$ and $P[1..] = \texttt{ACGTACT}$.

In the last case, we have a non-singleton block with suffixes of several segments, possibly with multiple occurences.
These suffixes represent the same substring of the original pangenome.
Here, we report a suffix array entry for each occurence of the substring in the pangenome.
To identify which entry is the first, we need to find the starting position with the smallest right context rank.
Because the ranks are sorted, this procedure is similar to the merging phase of a merge sort.
To iterate through all suffix array entries in the block, we therefore always identify the segment start with the next smallest right context rank and apply Equation \ref{eq:sa_entry} to this segment start.

As an example, consider the block at rows $[20..21]$ in the Suffix Table (\ref{tab:suffix}).
The relevant segment IDs are $0$ and $3$, with segment starts at positions $9$, $8$, $14$ and $0$.
The right context ranks from smallest to highest are $4, 5, 6, 9$ with the corresponding segment starts $8, 14, 0, 9$.
Applying Equation \ref{eq:sa_entry} to these segment starts yields a suffix array entries $8 + 0 = 8$, $14 + 0 = 14$, $0 + 0 = 0$ and $9 + 1 = 10$, representing pangenome suffixes $P[8..] = \texttt{CACACT}$, $P[14..] = \texttt{CACGACT}$, $P[0..] = \texttt{CACGTACT}$, and $P[10..] = \texttt{CACT}$.

\subsection{Application}
There are several proposed data structures implicitly relying on the structure
of prefix-free graphs.

>>>>

\section{Conclusion}
\subsection{Application}
There are several proposed data structures implicitly relying on the structure
of prefix-free graphs.

### 2023-06-24 manuscript/sa_construction.tex
finished sa construction

\subsection{Iteration}
With the previous tables stored in memory, we have all the necessary ingredients to generate the suffix array entries.

@@

% stoped here
Finaly, in the last case we report the positions.
The last case will partition the suffix table into the blocks of the same suffixes.
As an example consider the rows 20 and 21, which together for a single block.
All other blocks in the example consist of single rows and therefore we call them singletons.

This partitioning leads to three cases (from simplest to hardest):
\begin{itemize}
    \item A singleton block, occuring only once in the whole pangenome
    \item A singleton block, repeating several times in the pangenome
    \item A non-singleton block
\end{itemize}

In the case of a singleton block with only a single occurence we need to report only a single suffix array entry.
Given the row index $i$, this entry can be calculated with the following equation:

\begin{equation}
    \label{eq:sa_entry}
    \texttt{SA entry} = \texttt{starts}[\texttt{ID}[i]] + \texttt{pos}[i]
\end{equation}

As an example, consider the row $13$ in the Suffix Table (\ref{tab:suffix}), the first row yielding a SA value.
It's segment ID is $0$ and from the Segment Table (\ref{tab:segment}) we see that there is only one occurence of segment $0$ and it starts at position $9$ in the pangenome.
The offset from the start of a segment $\texttt{pos}[13]$ is $0$.
This yields the first entry of a suffix array $9 + 0 = 9$ corresponding to the lexicographically smallest suffix $P[9..] = \texttt{ACACT}$.

The second, slightly more complex case is a singleton block with several occurences in the pangenome.
For this case, we need to report as many suffix array entries as the number of occurences.
Because the starting segment positions in the Segment Table (\ref{tab:segment}) are sorted based on their right context rank, we can iterate through these starting positions and apply Equation \ref{eq:sa_entry} to each of them.

As an example, consider the row $14$ in the Suffix Table (\ref{tab:suffix}).
This suffix occurs twice in the pangenome in segments starting at positions $15$ and $1$.
Since the offset from the start of a segment $\texttt{pos}[14]$ is $0$, we report a suffix array entry $15 + 0 = 15$ and $1 + 0 = 1$, corresponding to the suffixes $P[15..] = \texttt{ACGACT}$ and $P[1..] = \texttt{ACGTACT}$.

In the last case, we have a non-singleton block with suffixes of several segments, possibly with multiple occurences.
These suffixes represent the same substring of the original pangenome.
Here, we report a suffix array entry for each occurence of the substring in the pangenome.
To identify which entry is the first, we need to find the starting position with the smallest right context rank.
Because the ranks are sorted, this procedure is similar to the merging phase of a merge sort.
To iterate through all suffix array entries in the block, we therefore always identify the segment start with the next smallest right context rank and apply Equation \ref{eq:sa_entry} to this segment start.

As an example, consider the block at rows $[20..21]$ in the Suffix Table (\ref{tab:suffix}).
The relevant segment IDs are $0$ and $3$, with segment starts at positions $9$, $8$, $14$ and $0$.
The right context ranks from smallest to highest are $4, 5, 6, 9$ with the corresponding segment starts $8, 14, 0, 9$.
Applying Equation \ref{eq:sa_entry} to these segment starts yields a suffix array entries $8 + 0 = 8$, $14 + 0 = 14$, $0 + 0 = 0$ and $9 + 1 = 10$, representing pangenome suffixes $P[8..] = \texttt{CACACT}$, $P[14..] = \texttt{CACGACT}$, $P[0..] = \texttt{CACGTACT}$, and $P[10..] = \texttt{CACT}$.

\subsection{Application}
There are several proposed data structures implicitly relying on the structure
of prefix-free graphs.

Rpair: rescaling repair with rsync \cite{2019gagie}
Prefix-free parsing for building big BWTs \cite{2019boucher}
MONI and r-index \cite{rossi2022moni,gagie2020fully}
Prefix-free Wheeler graphs \cite{2022pfwg}
MARIA \cite{2022maria}

All of these works implement their own version of prefix-free graphs.
We argue that separating prefix-free graphs as a standalone data structure can
bring several benefits:

\begin{itemize}
    \item reduce the complexity of the presentation
    \item allowing for better optimization of algorithms related to prefix-free graphs
    \item supporting the theoretical research by clearly deliminating the relevant terms
    \item allowing to bring prefix-free graphs closer to the biological data
\end{itemize}

Related to the last point, we remark that a possible future research can be in
the proper choice of trigger words.
As an example, we used stop codons in our experiments.
This could create prefix-free graphs capturing some biologically relevant phenomena.
Other options are recombination hotspots, highly repetitive elements, different
binding sites and sites with increased breakage.

>>>>

\subsection{Iteration}
With the previous tables stored in memory, we have all the necessary ingredients to generate the suffix array values.

@@

Finally, in the last case, we report the suffix array values.
The suffix table can be partitioned into blocks of the same segment suffixes.
For example, consider rows 20 and 21, which form a single block.
All other blocks in the running example consist of single rows; therefore, we call them singletons.

This partitioning leads to three cases:
\begin{itemize}
    \item a singleton block with segment suffix occurring only once in the whole pangenome
    \item a singleton block with segment suffix occurring several times in the pangenome
    \item a non-singleton block
\end{itemize}

In the first case, we must report only a single suffix array value.
Given the row index $i$, this value can be calculated with Equation \ref{eq:sa_value}.

\begin{equation}
    \label{eq:sa_value}
    \texttt{SA value} = \texttt{starts}[\texttt{ID}[i]] + \texttt{pos}[i]
\end{equation}

As an example, consider the row $13$ in the Suffix Table (\ref{tab:suffix}), the first row yielding a SA value.
Its segment ID is $0$, and from the Segment Table (\ref{tab:segment}), we see only one occurrence of segment $0$ with starting position $9$ in the pangenome.
The offset from the start of a segment $\texttt{pos}[13]$ is $0$.
Summing these two values, we get the first value of a suffix array $9 + 0 = 9$ corresponding to the lexicographically smallest suffix $P[9..] = \texttt{ACACT}$.

The second, slightly more complex case is a singleton block representing a segment suffix with several occurrences in the pangenome.
In this case, we must report as many suffix array values as the number of occurrences.
Because the starting segment positions in the Segment Table (\ref{tab:segment}) are sorted based on their right context rank, we can iterate through these starting positions and apply Equation \ref{eq:sa_value} to each of them.

As an example, consider the row $14$ in the Suffix Table (\ref{tab:suffix}).
This suffix occurs twice in the pangenome in segments starting at positions $15$ and $1$.
Since the offset from the start of a segment $\texttt{pos}[14]$ is $0$, we report a suffix array values $15 + 0 = 15$ and $1 + 0 = 1$, corresponding to the suffixes $P[15..] = \texttt{ACGACT}$ and $P[1..] = \texttt{ACGTACT}$.

In the last case, we have a non-singleton block representing suffixes of several segments, possibly with multiple occurrences.
These suffixes represent identical substrings in the original pangenome.
Here, we report a suffix array value for each of the substrings.
To identify the first value, we must find the starting position with the smallest right context rank.
Because the ranks are sorted, this procedure is similar to the merging phase of a merge sort.
Therefore, to iterate through all suffix array values in the block, we always identify the segment start with the next smallest right context rank and apply Equation \ref{eq:sa_value} to this segment start.

As an example, consider the block of rows $[20..21]$ in the Suffix Table (\ref{tab:suffix}).
The relevant segment IDs are $0$ and $3$, with segment starts at positions $9$, $8$, $14$ and $0$.
The right context ranks from smallest to highest are $4, 5, 6, 9$ with the corresponding segment starts $8, 14, 0, 9$.
Applying Equation \ref{eq:sa_value} to these segment starts yields a suffix array values $8 + 0 = 8$, $14 + 0 = 14$, $0 + 0 = 0$ and $9 + 1 = 10$, representing pangenome suffixes $P[8..] = \texttt{CACACT}$, $P[14..] = \texttt{CACGACT}$, $P[0..] = \texttt{CACGTACT}$, and $P[10..] = \texttt{CACT}$.
